Generating Correlated Data for Omics Simulation

### 2025-01-22 manuscript/paper.qmd
Make D^2 consistent, add funding sources

---
title: Generating Correlated Data for Omics Simulation
author:
    - name: "Jianing Yang"
      orcid: 0000-0002-2048-9398
      affiliations:
        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania"
        - name: "Chronobiology and Sleep Institute, University of Pennsylvania"
    - name: "Gregory R. Grant"
      orcid: 0000-0002-0139-7658
      affiliations:
        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania"
    - name: "Thomas G. Brooks"
      orcid: 0000-0002-6980-0079
      affiliations:
        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania"
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  cache: true
format:
  html:
    toc: true
    toc-expand: true
bibliography: references.bib
---

## Abstract

@@

Below, we describe the three methods for selecting the components of the covariance matrix $\Sigma := D + P P^T$.

@@

Since the distribution of the $\lambda'_i$ does not have a known analytic solution, we approximate this situation with the spiked Wishart distribution.
The rank $k-1$ Wishart distribution is that of the sample covariance matrix of $Y$ where the $n$ columns of $Y$ are iid $N(0, \Sigma)$.
The spiked Wishart is the special case where $\Sigma$ has $k$ arbitrary eigenvalues and the remaining are all equal to a constant.
Note that the singular values of $Y$ are the square roots of the eigenvalues of its sample covariance matrix.
While our case has $\Sigma$ non-diagonal, $\Sigma$ may be diagonalized by orthogonal rotations due the sepctral theorem, and orthogonal rotations do not change the singular values of $Y$.
Therefore, the distribution of singular values is not affected by the assumption that $\Sigma$ is diagonal.
Moreover, for the form $\Sigma = D + U W U^T$ where $p$ is very large, each column of $U$ is typically very close to orthogonal to any $e_i$, a standard basis vector.
Therefore, when $D = cI$ for some constant $c$, we can approximate $\Sigma$ as having $k$ arbitrary eigenvalues from $W$ and $n$ remaining eigenvalues all equal to $c$ corresponding to $D$.
This is a spiked Wishart distribution.

@@

1. Set $U$ to be the $p \times k$ matrix with columns $u_i$, the left singular vectors of $Z$.
2. Compute $w_1, \ldots, w_k$ and $c$ by stochastic gradient descent minimizing $\sum_{i} (E[\lambda'_i] - \lambda_i)^2$ for $\Sigma$ diagonal with entries $w_1, \ldots, w_k, c, \ldots, c$.
4. Set $W$ diagonal with the entries $w_1, \ldots, w_k$.
3. Set $D = cI$

@@

The `corpcor` package [@SchaferStrimmer2005; @OpgenRheinStrimmer2007] computes a James-Stein type shrinkage estimator for the covariance matrix.
For large $p$, this greatly improves the estimate of the covariance matrix by introducing a little bias towards zero correlations and equal variances of genes.
It computes optimal values of $\lambda_1$, and $\lambda_2$, its two regularization coefficients.
It then uses $\lambda_1$ to linearly interpolate the sample covariance matrix towards the identity matrix $I$ and $\lambda_2$ to interpolate the vector of variances towards the median variance value.
Since the sample covariance matrix is rank at most $n$, we again obtain a matrix of the form $\Sigma = D + UWU^T$.

@@

>>>>

---
title: Generating Correlated Data for Omics Simulation
author:
    - name: "Jianing Yang"
      orcid: 0000-0002-2048-9398
      affiliations:
        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania"
        - name: "Chronobiology and Sleep Institute, University of Pennsylvania"
    - name: "Gregory R. Grant"
      orcid: 0000-0002-0139-7658
      affiliations:
        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania"
    - name: "Thomas G. Brooks"
      orcid: 0000-0002-6980-0079
      affiliations:
        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania"
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  cache: true
format:
  html:
    toc: true
    toc-expand: true
bibliography: references.bib
---
## Abstract

@@

Below, we describe the three methods for selecting the components of the covariance matrix $\Sigma := D^2 + P P^T$.

@@

Since the distribution of the $\lambda'_i$ does not have a known analytic solution, we approximate this situation with the spiked Wishart distribution.
The rank $k-1$ Wishart distribution is that of the sample covariance matrix of $Y$ where the $n$ columns of $Y$ are iid $N(0, \Sigma)$.
The spiked Wishart is the special case where $\Sigma$ has $k$ arbitrary eigenvalues and the remaining are all equal to a constant.
Note that the singular values of $Y$ are the square roots of the eigenvalues of its sample covariance matrix.
While our case has $\Sigma$ non-diagonal, $\Sigma$ may be diagonalized by orthogonal rotations due the sepctral theorem, and orthogonal rotations do not change the singular values of $Y$.
Therefore, the distribution of singular values is not affected by the assumption that $\Sigma$ is diagonal.
Moreover, for the form $\Sigma = D^2 + U W U^T$ where $p$ is very large, each column of $U$ is typically very close to orthogonal to any $e_i$, a standard basis vector.
Therefore, when $D = c^2I$ for some constant $c$, we can approximate $\Sigma$ as having $k$ arbitrary eigenvalues from $W$ and $n$ remaining eigenvalues all equal to $c$ corresponding to $D$.
This is a spiked Wishart distribution.

@@

1. Set $U$ to be the $p \times k$ matrix with columns $u_i$, the left singular vectors of $Z$.
2. Compute $w_1, \ldots, w_k$ and $c$ by stochastic gradient descent minimizing $\sum_{i} (E[\lambda'_i] - \lambda_i)^2$ for $\Sigma$ diagonal with entries $w_1, \ldots, w_k, c^2, \ldots, c^2$ [@wishart].
4. Set $W$ diagonal with the entries $w_1, \ldots, w_k$.
3. Set $D = cI$

@@

The `corpcor` package [@SchaferStrimmer2005; @OpgenRheinStrimmer2007] computes a James-Stein type shrinkage estimator for the covariance matrix.
For large $p$, this greatly improves the estimate of the covariance matrix by introducing a little bias towards zero correlations and equal variances of genes.
It computes optimal values of $\lambda_1$, and $\lambda_2$, its two regularization coefficients.
It then uses $\lambda_1$ to linearly interpolate the sample covariance matrix towards the identity matrix $I$ and $\lambda_2$ to interpolate the vector of variances towards the median variance value.
Since the sample covariance matrix is rank at most $n$, we again obtain a matrix of the form $\Sigma = D^2 + UWU^T$.

@@

## Funding statement and competing interests

TB and GG received funding support from the National Center for Advancing Translational Sciences Grant (5UL1TR000003).
The funders had no role in this research, the decision to publish, or the preparation of this manuscript.

The authors declare no competing interests.
